---
title: "Data Wrangling with Tidyverse"
author: "Amanda Franklin"
date: "27/04/2022"
output: html_document
self_contained: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

## Introduction



## Setup

First we need to load packages and the dataset. If you are working in the R
project, the below code should work with no errors. If you aren't, you will need
to change the code to refer to the correct directory with the data.

```{r load packages, message=FALSE}
library(tidyverse)
dat <- read.csv("../data/datawrangling/datawrangling1.csv")
```

Now let's look at the dataset

```{r data1}
head(dat)
str(dat)

```

Oooof. It is not tidy. Let's get to cleaning it up.

## Edit Column Names

You will notice some of the column names have extra spaces at the end (you can
tell from the extra full stops). Use `rename()` to remove the extra full stop
from `Meral.Measured.` and `White.included.`

```{r edit column names}
dat <- dat %>%
  rename(Meral.Measured = Meral.Measured., White.included = White.included.)
colnames(dat) # Woohoo! Fixed
```

## Edit cells

Even in the first 6 lines of code (from the `head()` command) you may have 
noticed some inconsistencies. For example, `Meral.Measured` includes `y` and 
`Y` and the `Colour` column might not have consistent variable names.  
  
Let's check!

```{r data2}
dat %>% 
  dplyr::select(Meral.Measured, Colour) %>%  # Select the columns of interest
  mutate_all(as.factor) %>%  # Convert to factor so we can see the levels
  map(levels) # View levels for each column

```
Boooooo Amanda. Very inconsistent. Your task is to use `mutate` and `case_when()` 
to change the entries in the `Colour` column to `Green`, `Sandy` or `Brown`. For
this exercise, put `Browny/green` as `NA`.

Also use `mutate` and `toupper` and `str_trim` to clean up `Meral.Measured`

```{r edit cells}

dat2 <- dat %>%
  mutate(Meral.Measured = toupper(str_trim(Meral.Measured)),
         Colour = case_when(
           Colour == "Brown white dots" ~ "Brown",
           Colour == "sandy" ~ "Sandy",
           Colour == "Green/crème speckled" | 
             Colour == "Green with some white speckles" |
             Colour == "green" ~ "Green",
           Colour == "Browny/green" ~ NA_character_,
           TRUE ~ as.character(Colour)
         ))


# Let's Check

dat2 %>% 
  dplyr::select(Meral.Measured, Colour) %>%  # Select the columns of interest
  mutate_all(as.factor) %>%  # Convert to factor so we can see the levels
  map(levels) 

```

Yay! Ok let's move on to some other Tidyverse commands.

## Summarise data

Often you may want to quickly look or create a summary of your dataframe. For 
example, you  might want to calculate means and standard deviations for each
treatment or group in your dataset. This is easy with `group_by` and 
`summarise`.

So, let's give it a go! Calculate: 1) mean length, 2) sd (length) and 3) number 
of observations for each colour group.   
  
Extra task - remove the NAs from the grouping variable so they aren't included
in your summary table.

```{r summarise data}

summarytable <- dat2 %>%
  filter(!is.na(Colour)) %>%
  group_by(Colour) %>%
  summarise(AvgLength = mean(Length, na.rm = T),
            SDLength = sd(Length, na.rm = T),
            NLength = length(Length))
summarytable

```


## Long vs Wide Dataset

Sometimes, we have our dataset in long format and we need it in wide format, or
vice-versa. What do I mean by this? Imagine you have taken multiple measurements
from one individual on consecutive days. You could have one row per individual, 
and each measurement in separate columns like this:

```{r wide data}

dat_wide <- read.csv("../data/datawrangling/datawrangling2.csv")
head(dat_wide)

```

This is known as wide data. Long data would be if you had multiple rows for each
individual, a column indicating day number and then a column indicating the 
value for each day.

Your task is to turn the above dataframe into a long dataframe using 
`pivot_longer`. And then turn it back to a wide dataframe used `pivot_wider`

```{r pivot longer}

dat_long <- dat_wide %>%
  pivot_longer(!c(ID, Sex), names_to = "Day Number", values_to = "Values")
head(dat_long)

dat_wide2 <- dat_long %>%
  pivot_wider(names_from = `Day Number`, values_from = Values)
head(dat_wide2)

all_equal(dat_wide, dat_wide2)

```

## Merge Dataframes

Sometimes you might have two dataframes that you need to combine together. For
example, maybe you collected two different types of data (e.g. spec data and
behavioural data) and have these data in two different spreadsheets. If this is
the case, you just need a unique identifier to merge the two dataframes 
together.

In our case, we have ID in both `dat` and `dat_long`. Your task is to merge
together these two datasets using `inner_join`. This will associate the length,
widtch and colour data with the values in the data_long dataset. You will
end up with 3 rows for each ID.

```{r merging}

alldat <- dat %>%
  inner_join(., dat_long, by = "ID")
head(alldat)

```

